---
title: "Malaria Visualizations"
author: "James Lee"
date: "19/08/2022"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

Malaria is a mosquito-borne infectious disease caused by a parasite that affects humans. While the disease is uncommon in temperate climates, malaria is still common in tropical and subtropical countries. [Source: mayoclinic]


```{r}
# Import libraries & Set themes
library(malariaAtlas) # package specific to the Malaria Atlas Project
library(tidyverse) # data analytical package
theme_set(theme_light())
```

### 1. Malaria incidence in a specific country: Kenya Case Study 
```{r}
kenya_pr <- tbl_df(malariaAtlas::getPR(ISO = "KEN", species = "BOTH")) %>% # try Kenya
  filter(!is.na(pr)) # remove those w/o pr data
```

```{r}
# # 1 Plot pr over time
# kenya_pr %>%
#   group_by(year_start) %>%
#   summarise(examined = sum(examined),
#             positive = sum(positive),
#             studies = n()) %>%
#   mutate(pr = positive / examined) %>%
#   ggplot(aes(year_start, pr)) +
#   geom_line() +
#   geom_point()

# 2 Plot pr points on a 'map' in a specific point in time
kenya_pr %>%
  mutate(decade = 10 * (year_start %/% 10)) %>% # split the data by decade
  arrange(pr) %>% # this plots the highest pr last and highlights them over the lower pr pts for emphasis
  ggplot(aes(longitude, latitude, color = pr)) + 
  borders("world", regions = "Kenya") + # Superimpose these pr points on the Kenya map
  geom_point() +
  scale_color_gradient2(low = "blue", high = "red", midpoint = 0.5, labels = scales::percent_format()) +
  facet_wrap(~ decade) +
  theme_void() + # remove grids & axes
  coord_map() + # corrects map dim
  labs(title = "Malaria Prevalence in Kenya from 1980 to 2010", color = "Prevalence") 
```

Note that we use Kenya as a starting country as it was first in the data list, and also because it has a decent sample size to show any reliable trends. 

#### Key Insights
It is clearly evident that Malaria prevalence is concentrated in particular regions, namely the South & West of Kenya. 

It is noted over the decades, the high prevalence of Malaria always seem to occur in the same areas - the South & West of Kenya (see cluster of red points).

Over the decades, there are more Malaria screenings carried out in recent years (see increasing number of blue points over the decades), possibly attributed to Kenya's greater access to healthcare screenings.



### 2. Malaria incidence rates around the world

```{r}
malaria_inc <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_inc.csv")

malaria_inc_processed <- malaria_inc %>%
  setNames(c("country", "code", "year", "incidence")) %>% # mainly renames "entity" to "country" for readability
  mutate(incidence = incidence / 1000) # convert incidence data to percentages for readability
```

The plot below depicts the change in Malaria incidence in various countries in the year 2015 as compared to the year 2000. 

The change is calculated as the ratio of Malaria incidence in year 2015 over Malaria incidence in year 2000:

$$
	change=\frac{\text{# Malaria incidence in 2015}}{\text{# Malaria incidence in 2000}}
$$


```{r}
malaria_inc_processed %>%
  group_by(country) %>%
  summarize(current = last(incidence),
            change = last(incidence) / first(incidence)) %>% # we choose to calculate change in incidence via ratios
  arrange(desc(change))
```

```{r}
malaria_spread <- malaria_inc_processed %>%
  mutate(year = paste0("Y", year)) %>% # we do this so easier to mutate & calc change since years are numbers
  spread(year, incidence) # apply changes to all column names
  
malaria_spread %>%
  filter(country != "Turkey",
         !is.na(code)) %>%
  mutate(current = Y2015,
         change = Y2015 - Y2000) %>%
  # arrange(change)
  ggplot(aes(current, change)) +
  geom_point() +
  geom_text(aes(label = code), vjust = 1, hjust = 0)
```

Note that Turkey seemed to have 1741 incidences per 1000 people in year 2000, which does not seem logical. Hence, the plot above excludes data points from Turkey. Perhaps further work has to be done to determine the reason for this deviation.

#### Key Insights

The cluster of countries at the top-left of the plot indicates that most countries began with a small Malaria incidence had relative little change in Malaria incidence over time.

We clearly see that Venezuela was the only country with an increase in Malaria incidence from 2000, while all other countries exhibit varying levels of decrease in Malaria incidence.

Further, Mali currently (year 2015) has the highest Malaria incidence.

Lastly, Ethiopia had the largest decrease in Malaria incidence.



### 3. Malaria incidence evolution around the world

```{r}
world <- map_data("world") %>%
  filter(region != "Antarctica") # remove Antarctica

malaria_inc_processed %>%
  filter(incidence < 1) %>% # removes Turkey incidence
  inner_join(maps::iso3166 %>% # get country codes from iso3166 
             select(a3, mapname), by = c(code = "a3")) %>% #a3 refers to the country codes column in maps::is03166
  inner_join(world, by = c(mapname = "region")) %>% # keep all the borders in the world
  # inner join: focus on areas where we have data

# Now we can make a graph of the world
  ggplot(aes(long, lat, group = group, fill = incidence)) + # fill countries w pr
  geom_polygon() + # since we are dealing with filled maps
  scale_fill_gradient2(low = "blue", high = "red", midpoint = 0.2, labels = scales::percent_format()) +
  theme_void() +
  coord_map() +
  facet_wrap(~ year) +
  labs(title = "Malaria incidence over time around the world")

```

#### Key Insights

Notice that the Malaria incidence generally decreases over time, with most incidence occurring in Africa.



<!-- ### Malaria deaths over time around the world -->

<!-- ```{r} -->
<!-- malaria_deaths <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_deaths.csv") -->

<!-- malaria_deaths_processed <- malaria_deaths %>% -->
<!--   setNames(c("country", "code", "year", "deaths")) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # We are trying to join/match the data with its borders (ie join the two data sets tgt) -->

<!-- world <- map_data("world") %>% -->
<!--   filter(region != "Antarctica") # remove Antarctica -->

<!-- malaria_deaths_processed %>% -->
<!--   filter(year %in% c(2000, 2005, 2010)) %>% -->
<!--   inner_join(maps::iso3166 %>% -->
<!--              select(a3, mapname), by = c(code = "a3")) %>% -->
<!--   inner_join(world, by = c(mapname = "region")) %>% # keep all the borders in the world -->
<!--   # inner join: focus on areas where we have data -->

<!-- # Now we can make a graph of the world -->
<!--   ggplot(aes(long, lat, group = group, fill = deaths)) + # fill countries w pr -->
<!--   geom_polygon() + -->
<!--   scale_fill_gradient2(low = "blue", high = "red", midpoint = 100) + -->
<!--   theme_void() + -->
<!--   coord_map(xlim=c(-180,180)) + -->
<!--   facet_wrap(~ year) + -->
<!--   labs(title = "Malaria deaths over time around the world") -->

<!-- ``` -->

<!-- #### Key Insights: -->

<!-- There does not seem to be a significant trend or change in Malaria deaths over time around the world. In certain regions of Africa, Malaria deaths increased, while in other regions, it decreased. -->





Session Info:
```{r}
sessionInfo()
```




